Short Paper

### 2021-05-20 inst/SpatialDVM_Manuscript/SpatialDVM_Manuscript.tex
intro to model based and background of FPBK.

\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}

@@

@@

\hypertarget{the-elsevier-article-class}{%
\section{The Elsevier article class}\label{the-elsevier-article-class}}

@@

Here are two sample references: (Dirac, 1953; Feynman and Vernon Jr.;
1963).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Dirac1953888}{}%
Dirac, P.A.M., 1953. The lorentz transformation and absolute time.
Physica 19, 888--896.
doi:\href{https://doi.org/10.1016/S0031-8914(53)80099-6}{10.1016/S0031-8914(53)80099-6}

\leavevmode\hypertarget{ref-Feynman1963118}{}%
Feynman, R.P., Vernon Jr., F.L., 1963. The theory of a general quantum
system interacting with a linear dissipative system. Annals of Physics
24, 118--173.
doi:\href{https://doi.org/10.1016/0003-4916(63)90068-X}{10.1016/0003-4916(63)90068-X}

>>>>

\setcounter{secnumdepth}{5}
% Pandoc toggle for numbering sections (defaults to be off)

@@

\usepackage{bm} \DeclareMathOperator{\var}{{var}} \DeclareMathOperator{\cov}{{cov}}

@@

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

\hypertarget{design-based-philosphy}{%
\subsection{Design-Based Philosphy}\label{design-based-philosphy}}

\hypertarget{model-based-philosphy}{%
\subsection{Model-Based Philosphy}\label{model-based-philosphy}}

On the other hand, model-based inference imposes additional assumptions
on the data with a potential to provide more precise estimates if the
additional assumptions hold. Instead of estimating true but unknown
parameters, the goal of model-based inference in the spatial context is
often \emph{prediction} of an unknown quantity. This is a fundamental
philosophical difference between sampling-based and model-based
approaches. Instead of \emph{estimating} an fixed unknown mean, we are
\emph{predicting} the value of the mean, a random variable. We know that
if we sampled all sites, we would have an exact prediction for the mean
of our one realized spatial surface, without any uncertainty. But, the
true mean of the spatial process that generated our realized data is
still not known, and, in the prediction context, we typically do not
care much about what value the mean of the underlying process takes.

Figure 1a. Data is fixed. In a finite population example, show a 3d
surface that can be generated by anything. If we repeatedly sample the
surface, then 95\% of all 95\% CIs will contain the true mean, which
never changes.

Figure 1b. Spatial process is fixed. In a finite population example,
show 10 3d surfaces that are generated from some model. If we repeatedly
generate the surface and obtain a sample, then 95\% of all 95\% PIs will
contain the realized means. The realized mean changes from surface to
surface and it's not necessarily the case that 95\% of all 95\% PIs will
contain the true, underlying mean.

\hypertarget{comparing-design-based-vs.-model-based}{%
\subsection{Comparing Design-Based
vs.~Model-Based}\label{comparing-design-based-vs.-model-based}}

There have been many comparisons between the two paradigms\ldots\ldots.

\hypertarget{spatially-balanced-design-and-analysis}{%
\subsection{Spatially Balanced Design and
Analysis}\label{spatially-balanced-design-and-analysis}}

\hypertarget{finite-population-block-kriging}{%
\subsection{Finite Population Block
Kriging}\label{finite-population-block-kriging}}

Finite Population Block Kriging (FPBK) is an alternative to
samipling-based methods (Ver Hoef, 2008). FPBK expands the
geostatistical kriging framework to the finite population setting.
Instead of relying on a specific sampling design, we assume the data
were produced by a spatial stochastic process with spatial parameters
that can be estimated.

Ver Hoef (2008) gives details on the theory of FPBK, but some of the
basic principles are summarized below. For a response variable
\(\mathbf{z}\) that can be measured on a finite number of \(N\) sites,
our goal is to predict some linear function of all of the sample units,
\(\tau(\mathbf{z}) = \mathbf{b}^\prime \mathbf{z}\), where
\(\mathbf{b}\) is a vector of weights. One common vector of weights is a
vector of 1's to predict the total abundance in the region.

Typically, however, we only have a sample of the \(N\) sites. Denoting
quantities that are part of the sampled sites with a subscript \emph{s}
and quantities that are part of the unsampled sites with a subscript
\emph{u},

\begin{equation}
\begin{pmatrix} \label{equation:Zmarginal}
    \mathbf{z}_s      \\
    \mathbf{z}_u
\end{pmatrix}
=
\begin{pmatrix}
  \mathbf{X}_s    \\
  \mathbf{X}_u
\end{pmatrix}
\bm{\beta} +
\begin{pmatrix}
\bm{\delta}_s    \\
\bm{\delta}_u
\end{pmatrix},
\end{equation} where \(\mathbf{X}_s\) and \(\mathbf{X}_u\) are the
design matrices for the sampled and unsampled sites, respectively, and
\(\bm{\delta}_s\) and \(\bm{\delta}_u\) are zero-mean random errors for
the sampled and unsampled sites. Denoting
\(\bm{\delta} \equiv [\bm{\delta}_s \,\, \bm{\delta}_u]'\), then we
assume that \(E(\bm{\delta})\) = \(\mathbf{0}\).

We also typically assume that there is spatial correlation in
\(\bm{\delta}\), which can be modeled using a covariance function. Many
common choices for this function assume that spatial covariance
decreases with increasing Euclidean distance between sites. The primary
function used throughout the simulations and applications of this
manuscript is the Exponential covariance function: the \(i,j^{th}\)
entry for \(\mathop{\mathrm{{var}}}(\bm{\delta})\) is \mbox{}
\begin{equation}
\mathop{\mathrm{{cov}}}(\delta_i, \delta_j) = \theta_3 + \theta_1\exp(-h_{i,j}/\theta_2), \label{equation:expcov}
\end{equation} where \(h_{i,j}\) is the distance between sites \(i\) and
\(j\), and \(\bm{\theta}\) is a vector of spatial covariance parameters
of the partial sill \(\theta_1\), the range \(\theta_2\), and the nugget
\(\theta_3\). However, any spatial covariance function could be used in
the place of the Exponential, including functions that allow for
anisotropy {[}pg. 80 - 93{]}(Chiles and Delfiner, 1999).

With the above model formulation, the Best Linear Unbiased Predictor
(BLUP) for \(\tau(\mathbf{b}'\mathbf{z})\) and its prediction variance
can be computed. While details of the derivation are in (Ver Hoef,
2008), we note here that the predictor and its variance are both
moment-based. Neither require a particular distribution for
\(\mathbf{z}\).

\textbf{Software Implementation}: This probably goes in a different spot
(right before simulations?), but putting it here while it's on my mind.

FPBK can be readily performed in \texttt{R} with the \texttt{sptotal}
package (Matt et al., 2020). We use \texttt{sptotal} for both the
simulation analysis and the application, estimating parameters with
Restricted Maximum Likelihood (REML).

@@

Here are two sample references: (\textbf{Feynman1963118?};
\textbf{Dirac1953888?}).

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-ChilesEtAl1999GeostatisticsModelingSpatial}{}%
Chiles, J.-P., Delfiner, P., 1999. Geostatistics: {Modeling Spatial
Uncertainty}. {John Wiley \& Sons}, New York.

\leavevmode\hypertarget{ref-sptotal2020}{}%
Matt, H., Jay, V.H., Bryce, F., 2020. Sptotal: Predicting totals and
weighted sums from spatial data.

\leavevmode\hypertarget{ref-VerHoef2008spatial}{}%
Ver Hoef, J.M., 2008. Spatial methods for plot-based sampling of
wildlife populations. Environmental and Ecological Statistics 15, 3--13.
